---
title: "Exploratory Analysis in R"
output: 
  html_document: 
    theme: readable
    fig_caption: yes
    number_sections: yes
    toc: yes
    code_folding: show
    df_print: paged
  html_notebook: 
    fig_caption: yes
    number_sections: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

# **Exercise:** 

> Download the customer data and read into R.

```{r echo=TRUE, eval=FALSE}
customer <- read.csv(file = "https://xiaoruizhu.github.io/Data-Mining-R/lecture/data/CustomerData.csv")
```
>
> 1. How many rows and columns of the dataset?
> 2. Print first few rows the dataset.
> 3. Obtain the summary statistics (Min, Median, Max, Mean and Std.) for *Age*, *EducationYears*, *HHIncome*, and *CreditDebt*.
> 4. Obtain the mean of *HHIncome* by *MaritalStatus*
> 5. Obtain a pivot table of *LoanDefault* vs. *JobCategory*. Which Job Category has the highest and lowerst loan default rate?

```{r echo=FALSE, eval=FALSE}
customer<- read.csv("CustomerData.csv")
dim(customer)
head(customer)
names(customer)
summary(customer)
aggregate(HHIncome~MaritalStatus, data = customer, mean)
mypivot= table(customer$LoanDefault, customer$JobCategory)
mypivot[2,]/table(customer$JobCategory)
```

[go to top](#header)
