---
title: "tidyverse: collection of R packages for EDA"
output: 
  html_document: 
    theme: readable
    fig_caption: yes
    number_sections: yes
    toc: yes
    code_folding: show
    df_print: paged
  html_notebook: 
    fig_caption: yes
    number_sections: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 200)
```

# **tidyverse**: collection of R packages for EDA
We introduce package **tidyverse**, and some basic functions in the sub-packages for EDA. For more details, please see https://www.tidyverse.org/. This section is based on [Dr. Bradley Boehmke](http://bradleyboehmke.github.io/)'s [short course](http://uc-r.github.io/r_bootcamp) for MSBA students at Lindner College of Business. The course materials can be downloaded from [here](https://www.dropbox.com/sh/2qv0a02l9ausnyh/AAD0cRwMz_qTU-w15TwDxgLSa?dl=1).

```{r eval=FALSE}
install.packages("tidyverse")
```

```{r warning=FALSE}
library(tidyverse)
```

## Data Manipulation with **dplyr**

### Filtering and Indexing
We introduce **dplyr** package with some very user-friendly functions for data manipulation. These functions are:

* `filter()`
* `select()`
* `arrange()`
* `rename()`
* `mutate()`

#### Filtering (Subsetting) data

Here I introduce 4 ways to get subsets of data that satisfy certain logical conditions: `subset()`, logical vectors, SQL, and `filter()`. These kind of operations are called filtering in Excel. Knowing any one of these well is enough. Do not worry about memorizing the syntax, you can always look them up.

Suppose we want to get the **observations that have Sepal.Length > 5 and Sepal.Width > 4**. We can use logical operators: !=  not equal to; ==  equal to; |  or; & and.

- Use subset function
```{r, eval=FALSE}
data(iris)
subset(x = iris, subset = Sepal.Length > 5 & Sepal.Width > 4)
```
You can omit the x = and subset = part
```{r eval=FALSE}
subset(iris, Sepal.Length > 5 & Sepal.Width > 4)
```

- Use logical vectors
```{r}
iris[(iris$Sepal.Length > 5 & iris$Sepal.Width > 4), ]
```

- Use SQL statement
```{r, eval=FALSE}
install.packages('sqldf')
library(sqldf)
sqldf('select * from iris where `Sepal.Length` > 5 and `Sepal.Width` > 4')
```
In earlier version of sqldf all dots(.) in variable names need to be changed to underscores(_). 

- `filter()` is a power function in package **dplyr** to perform fitering like Excel Filter.
```{r}
# filter by row observations
data(iris)
iris_filter <- filter(iris, Sepal.Length<=5 & Sepal.Width>3)
iris_filter2 <- filter(iris, Species=="setosa", Sepal.Width<=3 | Sepal.Width>=4)
```

#### Subsetting the Dataset: Random Sample

The following code random sample (without replacement) 90% of the original dataset and assgin them to a new variable *iris_sample*. 
```{r}
iris_sample <- iris[sample(x = nrow(iris), size = nrow(iris)*0.90),]
```

The `dplyr` package provides more convinient ways for generating random samples. You can take a fixed number of samples using `sample_n()` or a fraction using `sample_frac()` as follows
```{r, eval=FALSE}
install.packages('dplyr')
library(dplyr)
iris_sample <- sample_frac(iris, 0.9)
```

The `dplyr` package provides more convinient ways for generating random samples. You can take a fixed number of samples using `sample_n()` or a fraction using `sample_frac()` as follows
```{r, eval=FALSE}
install.packages('dplyr')
library(dplyr)
iris_sample <- sample_frac(iris, 0.9)
# using dplyr for logical subsetting
filter(iris, Sepal.Length> 5, Sepal.Width > 4)
```
I recommend you to go through the [`dplyr` tutorial](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) and [lubridate tutorial](http://vita.had.co.nz/papers/lubridate.html). They make common data manipulation tasks and dealing with time-date much easier in R.


#### Sorting
Sorting by one or more variables is a common operation that you can do with datasets. With RStudio version 0.99+, you can sort a dataset when viewing it by clicking column header. 

To do it with code, let's suppose that you would like to find the top 5 rows in `iris` dataset with largest `Sepal.Length`.

```{r}
iris[order(iris$Sepal.Length, decreasing = TRUE)[1:5], ] 
```

The syntax is cleaner with the `arrange()` function in the `dplyr` package:
```{r, eval=FALSE}
arrange(iris, desc(Sepal.Length))[1:5, ]
```

#### Select columns

If you want to select one or more variables of a data frame, there are two ways to do that. First is using indexing by "[]". Second is `select()` function in *dplyr*. For example, suppose we want to select variable "Sepal.Length":
```{r eval=FALSE}
iris[, "Sepal.Length"]
```
or alternatively select two variables: "Sepal.Length", "Sepal.Width"
```{r eval=FALSE}
iris[, c("Sepal.Length", "Sepal.Width")]
```

On the other hand, `select()` in *dplyr* package can be used to filter by column, i.e., selecting or dropping variables.
```{r}
# Keep the variable Sepal.Length, Sepal.Width
varname <- c("Sepal.Length", "Sepal.Width")
iris_select <- select(iris, varname)
# verify if we did correctly
names(iris_select)
# This is equivalent to 
iris_select <- iris[,varname]
```

What about dropping variables? 
```{r}
iris_select2 <- select(iris, -Sepal.Length, -Petal.Length, -Species)
names(iris_select2)
```

This is equivalent to
```{r}
varname <- c("Sepal.Length", "Petal.Length", "Species")
iris_select2 <- iris[,!names(iris) %in% varname]
names(iris_select2)
```

#### Exercise
It would be easier if you know the order of the variables that you want to drop or keep. Try to obtain *iris_select* and *iris_select2* by using "dataname[, "variable_index"]."

### Re-ordering columns and sorting rows
Sorting by one or more variables is a common operation that you can do with datasets. With RStudio version 0.99+, you can sort a dataset when viewing it by clicking column header. 

To do it with code, let's suppose that you would like to find the top 5 rows in `iris` dataset with largest `Sepal.Length`.

```{r}
iris[order(iris$Sepal.Length, decreasing = TRUE)[1:5], ] 
```

The syntax is cleaner with the `arrange()` function in the `dplyr` package:
```{r, eval=FALSE}
arrange(iris, desc(Sepal.Length))[1:5, ]
```

```{r}
# re-ordering the columns
iris_order <- select(iris, Species, Petal.Width, everything())
names(iris_order)
# sorting rows by particular variable
iris_sort<- arrange(iris, Sepal.Length)
# sorting by more than one variable
iris_sort2<- arrange(iris, Sepal.Length, Sepal.Width)
# descending order
iris_sort_desc<- arrange(iris, desc(Sepal.Length))
```

Note that missing values are always sorted at the end.

### Renaming variable
```{r}
iris_rename<- rename(iris, SL=Sepal.Length, SW=Sepal.Width)
names(iris_rename)
```

### Creating New Variables
```{r}
iris_newvar<- mutate(iris, Sepal.L_W=Sepal.Length/Sepal.Width)
names(iris_newvar)
```

#### Exercise
Try to obtain *iris_newvar* WITHOUT using `mutate()` function. (You may need multiple steps, so `mutate()` is very useful especially you need to create many new variables.)

### Missing data

#### Detect missing values

Recall the customer dataset we in previous exercise.
```{r}
customer <- read.csv("data/CustomerData.csv")
```

```{r}
# How many missing values are in customer dataset?
sum(is.na(customer))
# How many missing values are in each variable?
data.frame(num_missing=colSums(is.na(customer)))
```

> **Exercise:**
> How to get only those variables that contain missing values? Which variable has the most missing?

#### How to deal with missing values?
* Simply delete those observations(rows) that contains missing value.
* Impute missing values by certain statistics, predictions or random number from estimated distributions.

```{r}
# Simply delete rows with missings
clean_customer<- na.omit(customer)
nrow(clean_customer)
# Impute missing values by median
medHS<- median(customer$HouseholdSize, na.rm = T)
customer$HouseholdSize[is.na(customer$HouseholdSize)==T]<- medHS
```

## Data Visualization with **ggplot2**

ggplot2 is a plotting system for R, based on the grammar of graphics, which tries to take the good parts of base and lattice graphics and none of the bad parts. It takes care of many of the fiddly details that make plotting a hassle (like drawing legends) as well as providing a powerful model of graphics that makes it easy to produce complex multi-layered graphics. More details can be found at http://ggplot2.org/. Here is a very good [tutorial](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html).

[go to top](#header)


## Things to remember

* How to obtain basic summary statistics
* Summary statistics by groups
* Pivot table
* Use of "[ ]" for subsetting and indexing
* Functions in `dplyr` packages.
* Basic R graphics.

[go to top](#header)