---
title: "Neural Networks Models"
header-includes:
   - \usepackage{float}
output: 
  html_document: 
    theme: readable
    fig_caption: yes
    number_sections: yes
    toc: yes
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```
```{r library, eval=TRUE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(xtable)
library(glmnet)
```

# Objective

This tutorial helps you to review various supervised learning techniques, introduce GAM, Neural Networks models, etc.

# Neural Networks Models
Neural Networks method (in-sample and out-of-sample performance measure) is illustrated here. The package [**nnet**](http://cran.r-project.org/web/packages/nnet/nnet.pdf) is used for this purpose.

## Regression

__Note__: 

- For regression problems add lineout = TRUE when training model. In addition, the response needs to be standardized to $[0, 1]$ interval. It's important normalize the response. If not, most of the times the algorithm will not converge. I chose to use the min-max method and scale the data in the interval [0,1]. 

```{r, message=FALSE}
library(MASS)
maxs <- apply(Boston, 2, max) 
mins <- apply(Boston, 2, min)

scaled <- as.data.frame(scale(Boston, center = mins, scale = maxs - mins))
index <- sample(1:nrow(Boston),round(0.9*nrow(Boston)))

train_Boston <- scaled[index,]
test_Boston <- scaled[-index,]

library(neuralnet)
```

```{r, fig.width=7}
n <- names(train_Boston)
f <- as.formula(paste("medv ~", paste(n[!n %in% "medv"], collapse = " + ")))
nn <- neuralnet(f,data=train_Boston, hidden=c(5,3), linear.output=T)
plot(nn)
```

```{r}
pr_nn <- compute(nn, test_Boston[,1:13])

pr_nn_org <- pr_nn$net.result*(max(Boston$medv)-min(Boston$medv))+min(Boston$medv)
test_r <- (test_Boston$medv)*(max(Boston$medv)-min(Boston$medv))+min(Boston$medv)
# MSE of testing set
MSE_nn <- sum((test_r - pr_nn_org)^2)/nrow(test_Boston)
MSE_nn
```

<!-- ## Classification -->

<!-- __Note__:  -->

<!-- - For classification problems with nnet you need to code the response to _factor_ first. In addition you want to add type = "class" for _predict()_  function.  -->

<!-- ### Credit Scoring Data -->

<!-- ```{r} -->
<!-- credit_data <- read.csv(file = "https://xiaoruizhu.github.io/Data-Mining-R/lecture/data/credit_default.csv", header=T) -->

<!-- # rename -->
<!-- library(dplyr) -->
<!-- credit_data<- rename(credit_data, default=default.payment.next.month) -->
<!-- # convert categorical data to factor -->
<!-- credit_data$SEX<- as.factor(credit_data$SEX) -->
<!-- credit_data$EDUCATION<- as.factor(credit_data$EDUCATION) -->
<!-- credit_data$MARRIAGE<- as.factor(credit_data$MARRIAGE) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- index <- sample(nrow(credit_data),nrow(credit_data)*0.90) -->
<!-- credit_train = credit_data[index,] -->
<!-- credit_test = credit_data[-index,] -->
<!-- ``` -->


<!-- You are already familiar with the credit scoring set. Let's define a cost function for benchmarking testing set performance. Note this is slightly different from the one we used for searching for optimal cut-off probability in logistic regression. Here the 2nd argument is the predict class instead of the predict probability (since many methods are not based on predict probability). -->

<!-- ```{r} -->
<!-- creditcost <- function(observed, predicted){ -->
<!--   weight1 <- 5 -->
<!--   weight0 <- 1 -->
<!--   c1 <- (observed==1)&(predicted == 0) #logical vector - true if actual 1 but predict 0 -->
<!--   c0 <- (observed==0)&(predicted == 1) #logical vector - true if actual 0 but predict 1 -->
<!--   return(mean(weight1*c1+weight0*c0)) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r, message=FALSE} -->
<!-- library(nnet) -->
<!-- credit_nnet <- nnet(default~., data=credit_train, size=20, maxit=5000) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- prob_nnet <- predict(credit_nnet, credit_test) -->
<!-- pred_nnet <- as.numeric(prob_nnet > 0.05) -->
<!-- table(credit_test$default,pred_nnet, dnn=c("Observed","Predicted")) -->
<!-- cat("Misclassification rate =", mean(ifelse(credit_test$default != pred_nnet, 1, 0))) -->
<!-- cat("Asymmetric cost (with ratio 10:1) =", creditcost(credit_test$default, pred_nnet)) -->
<!-- ``` -->

[go to top](#header)
