---
title: "Performance Comparisons"
header-includes:
   - \usepackage{float}
output: 
  html_document: 
    theme: readable
    fig_caption: yes
    number_sections: yes
    toc: yes
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```
```{r library, eval=TRUE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(xtable)
library(glmnet)
```

# Performance Comparisons

At last, after fitting several models, you may want to compare their in-sample and out-of-sample performances. The performance measures are illustrated in previous sections. In your report, you may want to put them in some table format. Note that not all measures are applicable. For example, I didn't find AIC/BIC or deviance for LDA models and Neural Network models. For tree models, *tree* package can give you mean residual deviance but not with *rpart* package. If you find either one of them, I would be interested to know.

## In-sample
You may compare the following
- AIC or BIC
- Mean Residual Deviance (for binary response) or Mean Square Error (for continuous response)
- Cost (asymmetric or symmetric)
- Misclassification Rate
- ROC curve or Area Under the Curve (AUC)

## Out-of-sample
- Cost
- Misclassification Rate
- ROC curve or Area Under the Curve (AUC)


## Symmetric Cost and Multiclass Problems

For classification tasks with symmetric costs many of functions can be simplified. You do not have to worry about the cut-off probability and can focus on the tuning parameters in each model (e.g. cost and gamma in SVM).

Different classifiers deal with multiclass classification differently. Logistic regression can be extended to multinomial logistic regression (using _multinom_ function). Many other binary classifiers can use an either "one-vs-all"(train N binary classifiers to distinguish each class from the rest) or "one-vs-one"(train C(N,2) binary classifiers for each possible pair of classes) approach to deal with multiple classes. 

```{r}
library(e1071)
data(iris)
id_train <- sample(nrow(iris),nrow(iris)*0.80)
iris.train = iris[id_train,]
iris.test = iris[-id_train,]
iris.svm = svm(Species ~ ., data = iris.train)
table(iris.test$Species, predict(iris.svm, iris.test), dnn=c("Observed","Predicted"))
```

